Reference label transfer of the fetal kidney reference (Stewart et al.) for `r params$sample_id`

### 2024-08-16 analyses/cell-type-wilms-tumor-06/notebook_template/02a_label-transfer_fetal_full_reference_Cao.Rmd
changes to #706

---
title: "Reference label transfer exploration for `r params$sample_id`"
author: "Maud PLASCHKA"
date: '2024-08-07'
params:
  current: "2024-07-08"
  scpca_project_id: "SCPCP000006"
  sample_id: "SCPCS000176"
  seed: 12345
output: 
  html_document: 
    toc: yes
    toc_float: yes
    code_folding: hide
    highlight: pygments
    df_print: paged
---

@@

The aim is to find an appropriate reference to perform label transfer as a step of the annotation of the Wilms tumor dataset for the sample `param$sample_id` =

@@

We thus decided to test and compare fetal (kidney) references that could be use in the analysis module.
You can find below the three workflow/reference that we plan to test.

@@

We first wanted to try the human fetal kidney atlas to transfer label into the Wilms tumor samples using azimuth. 
You can find more about the human kidney atlas here: https://www.kidneycellatlas.org/ [3]

Using this reference, we test two workflows for label transfer:

- `Azimuth v5`,
- `Seurat v4` as described here  https://satijalab.org/seurat/articles/integration_mapping.html#cell-type-classification-using-an-integrated-reference

@@

##########################################################################################################
In this notebook, we will test the human fetal Azimuth reference from Cao et al. using Azimuth . 
##########################################################################################################

@@

# The current data directory, found within the repository base directory
data_dir <- file.path(repository_base, "data", params$current, params$scpca_project_id)

@@

In this notebook, we test the label transfer from the azimuth fetal atlas (reference) to the Wilms tumor sample defined in `params$sample_id` from the Wilms tumor dataset `param$scpca_project_id`

```{r path_to_query}
filelist <- list.files(data_dir, full.names = TRUE)

# select the 40 Wilms tumor single nucleus folders only
filedir <- filelist[grepl(params$sample_id, filelist)]
```

@@

Reports will be saved in the `notebook/00-reference` directory.

### Load and pre-process the query data

```{r load, message=FALSE, warning=FALSE}
file <- list.files(filedir)
file <- file[grepl("_processed.rds", file)]

# open the processed rds object
sce <- readRDS(paste0(filedir, "/", file))
```

```{r seurat_wf, fig.height=4, fig.width=8, warning=FALSE, out.width='100%'}
# convert to seurat
srat <- CreateSeuratObject(counts = counts(sce),
                                    assay = "RNA",
                                    project = params$sample_id
                           )

# convert colData and rowData to data.frame for use in the Seurat object
cell_metadata <- as.data.frame(colData(sce))
row_metadata <- as.data.frame(rowData(sce))

# add cell metadata (colData) from SingleCellExperiment to Seurat
srat@meta.data <- cell_metadata

# add row metadata (rowData) from SingleCellExperiment to Seurat
srat[["RNA"]]@meta.data <- row_metadata

# add metadata from SingleCellExperiment to Seurat
srat@misc <- metadata(sce)

# Normalization
options(future.globals.maxSize= 8912896000000)
srat <- SCTransform(srat, verbose = F, conserve.memory = TRUE)

# dimensionality reduction
srat <- RunPCA(srat, verbose = F)
srat <- RunUMAP(srat, dims = 1:50, verbose = F)

# clustering
srat    <- FindNeighbors(srat, dims = 1:50, verbose = F)
srat    <- FindClusters(srat, verbose = T)

```

@@

```{r fig.height=6, fig.width=6, message=FALSE, warnings=FALSE}
DefaultAssay(srat) <- "RNA"
s <- RunAzimuth(srat, reference ="fetusref" )
```

```{r fig.height=15, fig.width=8, warnings=FALSE}

d1 <- DimPlot(s, reduction = "umap", dims = c(1,2), group.by = "predicted.organ", label = TRUE, repel = TRUE) + NoLegend()
d2 <- DimPlot(s, reduction = "umap", dims = c(1,2), group.by = "predicted.annotation.l1", label = TRUE, repel = TRUE) + NoLegend()
d3 <- DimPlot(s, reduction = "umap", dims = c(1,2), group.by = "predicted.annotation.l2", label = TRUE, repel = TRUE) + NoLegend()

f1 <- SCpubr::do_BarPlot(sample = s,
                         group.by = "predicted.organ",
                         split.by = "seurat_clusters",
                         position = "fill",
                         font.size = 10,
                         legend.ncol = 4) +
                         ggtitle("% cells")+
                         xlab(file)

f2 <- SCpubr::do_BarPlot(sample = s,
                         group.by = "predicted.annotation.l1",
                         split.by = "seurat_clusters",
                         position = "fill",
                         font.size = 10,
                         legend.ncol = 2) +
                         ggtitle("% cells")+
                         xlab(file)

f3 <- SCpubr::do_BarPlot(sample = s,
                         group.by = "predicted.annotation.l2",
                         split.by = "seurat_clusters",
                         position = "fill",
                         font.size = 10,
                         legend.ncol = 2) +
                         ggtitle("% cells")+
                         xlab(file)

@@

>>>>

---
title: "Reference label transfer from the fetal full reference (Cao et al.) for `r params$sample_id`"
author: "Maud PLASCHKA"
date: '2024-08-07'
params:
  scpca_project_id: "SCPCP000006"
  sample_id: "SCPCS000176"
  seed: 12345
output: 
  html_document: 
    toc: yes
    toc_float: yes
    code_folding: hide
    highlight: pygments
    df_print: paged
---

@@

The aim is to perform label transfer as a step of the annotation of the Wilms tumor dataset for the sample `param$sample_id` =

@@

We thus decided to test and compare fetal (kidney) references that could be use in the analysis module.
You can find below the two references that we plan to use

@@

Stewart et al. created and a human fetal kidney atlas.
We will use `Azimuth` to transfer label into the Wilms tumor samples. 
You can find more about the human kidney atlas here: https://www.kidneycellatlas.org/ [3]

@@

In this notebook, we perform label transfer from the human fetal Azimuth reference from Cao et al.

@@

@@

In this notebook, we test the label transfer from the azimuth fetal atlas (reference) to the Wilms tumor sample defined in `params$sample_id` from the Wilms tumor dataset `param$scpca_project_id`.

We work with the pre-processed `Seurat`object saved in the `results` directory.

```{r path_to_query}
data_dir <- file.path(module_base, "results", params$sample_id)
```

@@

Reports will be saved in the `notebook/02-reference` directory. 
The pre-processed and annotated `Seurat` object per samples are saved in the `result` folder.
```{r path_to_output}
output_dir <- file.path(module_base, "results", params$sample_id)
```

## Analysis

### Load the pre-process data (query)

```{r load, message=FALSE, warning=FALSE}
# open the processed rds object
srat <- readRDS(file.path(data_dir, paste0("01-Seurat_", params$sample_id,".Rds")))
```

@@

```{r run_azimuth, message=FALSE, warnings=FALSE}
DefaultAssay(srat) <- "RNA"
s <- RunAzimuth(srat, reference ="fetusref" )

# We transfer the annotation to the pre-processed `Seurat`iobject as we don't want to keep changes done on the query by `RunAzimuth`
metadata_to_trasfer <- s@meta.data[, c("predicted.annotation.l1.score", "predicted.annotation.l1", "predicted.annotation.l2.score", "predicted.annotation.l2", "predicted.organ.score", "predicted.organ")]
srat <- AddMetaData(srat, metadata_to_trasfer, col.name = c("fetal_full_predicted.annotation.l1.score", "fetal_full_predicted.annotation.l1", "fetal_full_predicted.annotation.l2.score", "fetal_full_predicted.annotation.l2", "fetal_full_predicted.organ.score", "fetal_full_predicted.organ"))
```

```{r plot_azimuth, fig.height=15, fig.width=8, warnings=FALSE}

d1 <- DimPlot(srat, reduction = "umap", dims = c(1,2), group.by = "fetal_full_predicted.organ", label = TRUE, repel = TRUE) + NoLegend()
d2 <- DimPlot(srat, reduction = "umap", dims = c(1,2), group.by = "fetal_full_predicted.annotation.l1", label = TRUE, repel = TRUE) + NoLegend()
d3 <- DimPlot(srat, reduction = "umap", dims = c(1,2), group.by = "fetal_full_predicted.annotation.l2", label = TRUE, repel = TRUE) + NoLegend()

f1 <- SCpubr::do_BarPlot(sample = srat,
                         group.by = "fetal_full_predicted.organ",
                         split.by = "seurat_clusters",
                         position = "fill",
                         font.size = 10,
                         legend.ncol = 4) +
                         ggtitle("% cells")+
                         xlab(file)

f2 <- SCpubr::do_BarPlot(sample = srat,
                         group.by = "fetal_full_predicted.annotation.l1",
                         split.by = "seurat_clusters",
                         position = "fill",
                         font.size = 10,
                         legend.ncol = 2) +
                         ggtitle("% cells")+
                         xlab(params$sample_id)

f3 <- SCpubr::do_BarPlot(sample = srat,
                         group.by = "fetal_full_predicted.annotation.l2",
                         split.by = "seurat_clusters",
                         position = "fill",
                         font.size = 10,
                         legend.ncol = 2) +
                         ggtitle("% cells")+
                         xlab(params$sample_id)

@@

## Save the `Seurat`object

```{r save}
saveRDS(object = srat, file = file.path(output_dir, paste0("00a-fetal_full_label-transfer_",params$sample_id,".Rds")))
```

### 2024-08-16 analyses/cell-type-wilms-tumor-06/notebook_template/02b_label-transfer_fetal_kidney_reference_Stewart.Rmd
changes to #706

---
title: "Reference label transfer exploration for `r params$sample_id`"
author: "Maud PLASCHKA"
date: '2024-08-13'
params:
  current: '2024-07-08'
  scpca_project_id: "SCPCP000006"
  sample_id: "SCPCS000176"
  ref_url: "https://datasets.cellxgene.cziscience.com/40ebb8e4-1a25-4a33-b8ff-02d1156e4e9b.rds"
  seed: 12345
output: 
  html_document: 
    toc: yes
    toc_float: yes
    code_folding: hide
    highlight: pygments
    df_print: paged
---

@@

##########################################################################################################
In this notebook, we will test the human fetal kidney atlas from Stewart et al. using the Azimuth workflow. 
##########################################################################################################

@@

# The current data directory, found within the repository base directory
data_dir <- file.path(repository_base, "data", params$current, params$scpca_project_id)

@@

rds data can be downloaded using the URL defined in `param$ref_url`.

Please note that this download link permanently references the current version of the dataset (08/2024). 
If this dataset is updated, a new download link will be created that permanently references the next version of this dataset.

We place it in the `scratch` folder of the module because it is a temporary file.

```{r path_to_reference}
path_to_reference <- file.path(module_base, "scratch", "fetal_full.rds")
```

@@

In this notebook, we test the label transfer from the fetal kidney atlas (reference) to the Wilms tumor sample defined in `params$sample_id` from the Wilms tumor dataset `param$scpca_project_id`

```{r path_to_query}
filelist <- list.files(data_dir, full.names = TRUE)

# select the 40 Wilms tumor single nucleus folders only
filedir <- filelist[grepl(params$sample_id, filelist)]
```

@@

This notebook has two main outputs

#### Azimuth compatible reference

The reference data are downloaded from the `params$ref_url` and transiently saved in the `scratch`folder.
The azimuth compatible fetal kidney reference will be saved in the `results` folder from the module, `reference` subfolder.

Please note that the creation of the Azimuth reference cannot be done within RMarkdown. 
We thus set `include = FALSE` to the corresponding chunk.
Please, before running this code either :

- download the azimuth compatible reference from S3 bucket or
- run manualy chunk by chunk the report to create and save the reference once.

```{r path_to_output}
path_to_output <- file.path(module_base, "results", "references")
```

#### The html notebook saved afetr rendering this notebook (knit).
Will be saved in the `notebook/00-reference` directory.

## Using the Human Kidney Atlas - Fetal as a reference

### Create a azimuth compatible reference

Note: not shown here because the `AzimuthReference` function do not allow knitr.
The reference should be build "manually" prior to knitr the report.

```{r create_ref, eval=FALSE, fig.height=7, fig.width=12, message=FALSE, warning=FALSE, include=FALSE, out.width='100%'}
download.file(params$ref_url, path_to_reference)
seurat <- readRDS(path_to_reference)

options(future.globals.maxSize= 891289600000)
s <- SCTransform(seurat, verbose = FALSE, method = "glmGamPoi", conserve.memory = TRUE)
s <- RunPCA(s, npcs = 50, verbose = FALSE)
s <- RunUMAP(s, dims = 1:50, verbose = FALSE, return.model = TRUE)

Fetal_kidney <- AzimuthReference(
  s,
  refUMAP = "umap",
  refDR = "pca",
  refAssay = "SCT",
  dims = 1:50,
  k.param = 31,
  plotref = "umap",
  plot.metadata = NULL,
  ori.index = NULL,
  colormap = NULL,
  assays = NULL,
  metadata = c("compartment", "cell_type"),
  reference.version = "0.0.0",
  verbose = FALSE
)

d2 <- do_DimPlot(Fetal_kidney, reduction = "refUMAP", dims = c(1,2), group.by = "compartment", label = TRUE, repel = TRUE) + NoLegend() + ggtitle("umap reduction after SCTransform and Seurat workflow")
d3 <- do_DimPlot(Fetal_kidney, reduction = "refUMAP", dims = c(1,2), group.by = "cell_type", label = TRUE, repel = TRUE) + NoLegend()

d2 | d3

if (file.exists(path_to_output)){
   SaveAnnoyIndex(object = Fetal_kidney[["refdr.annoy.neighbors"]], file = file.path(path_to_output, "idx.annoy"))
    saveRDS(object = Fetal_kidney, file = file.path(path_to_output, "ref.Rds"))
} else {
    SaveAnnoyIndex(object = Fetal_kidney[["refdr.annoy.neighbors"]], file = file.path(path_to_output, "idx.annoy"))
    saveRDS(object = Fetal_kidney, file = file.path(path_to_output, "ref.Rds"))

}

# finally, we can remove the fetal_full.rds file from the marker-sets directory
file.remove(path_to_reference)
```

### Load and pre-process the query data

```{r load, message=FALSE, warning=FALSE}
file <- list.files(filedir)
file <- file[grepl("_processed.rds", file)]

# open the processed rds object
sce <- readRDS(paste0(filedir, "/", file))
```

```{r seurat_wf, fig.height=4, fig.width=8, warning=FALSE, out.width='100%'}
# convert to seurat
srat <- CreateSeuratObject(counts = counts(sce),
                                    assay = "RNA",
                                    project = params$sample_id
                           )

# convert colData and rowData to data.frame for use in the Seurat object
cell_metadata <- as.data.frame(colData(sce))
row_metadata <- as.data.frame(rowData(sce))

# add cell metadata (colData) from SingleCellExperiment to Seurat
srat@meta.data <- cell_metadata

# add row metadata (rowData) from SingleCellExperiment to Seurat
srat[["RNA"]]@meta.data <- row_metadata

# add metadata from SingleCellExperiment to Seurat
srat@misc <- metadata(sce)

# Normalization
options(future.globals.maxSize= 8912896000000)
srat <- SCTransform(srat, verbose = F, conserve.memory = TRUE)

# dimensionality reduction
srat <- RunPCA(srat, verbose = F)
srat <- RunUMAP(srat, dims = 1:50, verbose = F)

# clustering
srat    <- FindNeighbors(srat, dims = 1:50, verbose = F)
srat    <- FindClusters(srat, verbose = T)

```

@@

```{r fig.height=6, fig.width=6, message=FALSE, warnings=FALSE}

data <-  RunAzimuth(srat, path_to_output, assay = 'RNA')

```

```{r fig.height=8, fig.width=8, warnings=FALSE}

d1 <- DimPlot(data, reduction = "umap", dims = c(1,2), group.by = "predicted.compartment", label = TRUE, repel = TRUE) + NoLegend()
d2 <- DimPlot(data, reduction = "umap", dims = c(1,2), group.by = "predicted.cell_type", label = TRUE, repel = TRUE) + NoLegend()

f1 <- SCpubr::do_BarPlot(sample = data,
                         group.by = "predicted.compartment",
                         split.by = "seurat_clusters",
                         position = "fill",
                         font.size = 10,
                         legend.ncol = 3) +
                         ggtitle("% cells")+
                         xlab(file)

f2 <- SCpubr::do_BarPlot(sample = data,
                         group.by = "predicted.cell_type",
                         split.by = "seurat_clusters",
                         position = "fill",
                         font.size = 10,
                         legend.ncol = 3) +
                         ggtitle("% cells")+
                         xlab(file)

@@

>>>>

---
title: "Reference label transfer of the fetal kidney reference (Stewart et al.) for `r params$sample_id`"
author: "Maud PLASCHKA"
date: '2024-08-13'
params:
  scpca_project_id: "SCPCP000006"
  sample_id: "SCPCS000176"
  seed: 12345
output: 
  html_document: 
    toc: yes
    toc_float: yes
    code_folding: hide
    highlight: pygments
    df_print: paged
---

@@

In this notebook, we perform label transfer from the human fetal kidney atlas from Stewart et al.

@@

# The current data directory, found within the repository base directory
data_dir <- file.path(repository_base, "data", "current", params$scpca_project_id)

@@

The reference has been download and pre-process in the `R Script` `download-and-create-fetal-kidney-ref.R`

```{r path_to_reference}
path_to_reference <- file.path(module_base, "results", "references")
```

@@

In this notebook, we test the label transfer from the fetal kidney atlas (reference) to the Wilms tumor sample defined in `params$sample_id` from the Wilms tumor dataset `param$scpca_project_id` 
We work with the pre-processed `Seurat`object and labeled with the fetal kidney reference saved in the `results` directory.

```{r path_to_query}
data_dir <- file.path(module_base, "results", params$sample_id)
```

@@

Reports will be saved in the `notebook/02-reference` directory. 
The pre-processed and annotated `Seurat` object per samples are saved in the `result` folder.

```{r path_to_output}
output_dir <- file.path(module_base, "results", params$sample_id)
```

## Analysis

### Load the pre-processed `Seurat` object
```{r load, message=FALSE, warning=FALSE}
# open the processed rds object
srat <- readRDS(file.path(data_dir, paste0("00a-fetal_full_label-transfer_",params$sample_id,".Rds")))
```

@@

```{r run_azimuth, message=FALSE, warnings=FALSE}

data <-  RunAzimuth(srat, path_to_reference, assay = 'RNA')
# We transfer the annotation to the pre-processed `Seurat`object as we don't want to keep changes done on the query by `RunAzimuth`
metadata_to_trasfer <- data@meta.data[, c("predicted.compartment.score", "predicted.compartment", "predicted.cell_type.score", "predicted.cell_type")]
srat <- AddMetaData(srat, metadata_to_trasfer, col.name = c("fetal_kidney_predicted.compartment.score", "fetal_kidney_predicted.compartment", "fetal_kidney_predicted.cell_type.score", "fetal_kidney_predicted.cell_type"))
```

```{r plot_azimuth, fig.height=8, fig.width=8, warnings=FALSE}

d1 <- DimPlot(srat, reduction = "umap", dims = c(1,2), group.by = "fetal_kidney_predicted.compartment", label = TRUE, repel = TRUE) + NoLegend()
d2 <- DimPlot(srat, reduction = "umap", dims = c(1,2), group.by = "fetal_kidney_predicted.cell_type", label = TRUE, repel = TRUE) + NoLegend()

f1 <- SCpubr::do_BarPlot(sample = srat,
                         group.by = "fetal_kidney_predicted.compartment",
                         split.by = "seurat_clusters",
                         position = "fill",
                         font.size = 10,
                         legend.ncol = 3) +
                         ggtitle("% cells")+
                         xlab(params$sample_id)

f2 <- SCpubr::do_BarPlot(sample = srat,
                         group.by = "fetal_kidney_predicted.cell_type",
                         split.by = "seurat_clusters",
                         position = "fill",
                         font.size = 10,
                         legend.ncol = 3) +
                         ggtitle("% cells")+
                         xlab(params$sample_id)

@@

## Save the `Seurat`object

```{r save}
saveRDS(object = srat, file = file.path(output_dir, paste0("00b-fetal_kidey_label-transfer_",params$sample_id,".Rds")))
```
